---
title: "Why Did President Barak Obama Win Again If His Approval Ratings Were Declining?"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

# 1. Introduction

```{r load libraries, message=FALSE, warning=FALSE, include=FALSE}
packages.used=as.list(
  c(
  "tidyverse",
  "haven",
  "devtools",
  "RColorBrewer",
  "data.table",
  "ggplot2",
  "dplyr",
  "knitr")
)

check.pkg = function(x){
  if(!require(x, character.only=T)) install.packages(x, 
                                                     character.only=T,
                                                     dependence=T)
}

lapply(packages.used, check.pkg)
```

```{r read in data, message=FALSE, include=FALSE}
setwd("~/ADS/Fall2020-Project1-sss2289/data")
library(haven)
anes_dat <- read_sav("anes_timeseries_2012.sav")
```


```{r echo=FALSE}
par(mar=c(10, 4,1,4))
#summary(anes_dat$prevote_presvtwho)
#str(anes_dat$prevote_presvtwho)
#levels(factor(anes_dat$prevote_presvtwho))
barplot(table(anes_dat$prevote_primvwho),names.arg=c("Refused to Answer","Do Not Know","Inapplicable","Mitt Romney","Barack Obama","Rick Santorum","Newt Gingrich","Ron Paul","Rick Perry","Michele Bachmann","Jon Huntsman ", "Herman Cain","Other Candidate"),las=2,main="2012 Primary Election", ylab ="count")
 mtext("Candidates' Names", side=1, line =8)
par(mar=c(5.1, 4.1, 4.1, 2.1))
```

The graph above shows the vote counts of 2012 primary elections and it includes 'inapplicable' counts bicide the individual candidates counts.

For now we will exclude the inapplicable (will discuss more about it later). This bar graph shows that a large number of individuals voted for Barack Obama over other candidates in the 2012 Primaries. From what I remember, 2012 Primaries were held in November to December range, during that time period Obama’s approval ratings were at their lowest. However, this graph seems to show that despite this people still preferred him to be the president over all others.

Now about the inapplicable bar: In some states if you are not affiliated to a specific party (you declare yourself as independent) then you cannot vote in the primaries. Also, the inapplicable column could refer to individuals that are not citizens age 18 and older. If you are not older than 17 years and you are not a citizen then you cannot vote in United States of America. Another explanation about the inapplicable column could be because people do not have an opinion about any of the candidates. Lastly, the inapplicable column could be because people got paid to take the survey, they did not put thought into selecting their answers truthfully. If the above-mentioned biases did not exist, then the count for people that voted in the primaries could have been higher. Further solidifying the fact that more people voted for Obama than other candidates or it could change which candidate won in the primaries. Ignoring the inapplicable answers, if you look at the people that did vote most of them voted for Obama. These results beg for further digging and form the basis of my question.


The 44th President and the First African American President, Barack Obama was in office for two terms. At the end of his first term his approval ratings fell from the high 60% is to about 40%. However, despite his declining approval ratings he managed to get elected for a second term in office. This begs the question, if President Barack Obama's approval ratings were declining what caused him to win again in 2012?


```{r warning=FALSE, include=FALSE}
#(duplicated(anes_dat$caseid)) check for duplicate entries
library("dplyr")
ncol(anes_dat)
nrow(anes_dat)
```

# 2. About Data
I have used 2012 Time Series Study Survey Data from the 'American National Election Studies' (ANES). For the first time ANES used the Internet to conduct it's survey. The participants that filled out the online survey were selected from the 'GfK Knowledge Networks'. Thus, the data was compiled from respondents that filled out the online administered survey and compiled from traditional survey method (a person asks the respondents questions). The participants were given money incentives to fill out the pre and post election surveys, amounts ranged from $65 to $125. The pre part of the survey was conducted from September 1st 2012 to November 5th 2012 and the post part of the survey was conducted from November 6th 2012 to January 1st 2013. In this data set there are, `r ncol(anes_dat)`, variables in total and, `r nrow(anes_dat)`, respondents have filled out this survey (Pre and Post). 


# 3. Methodology
During any presidential election there typically three big ticket subjects that the public looks at when selecting a candidate to vote for. The general population looks at what the future president stance is on health care, how they are going to handle economic problems, and how they are going to handle foreign relations with other countries. Hence, to answer the question I posed above, I looked at a variables from the survey that answered these 3 big ticket subjects.

# 4. Analysis

<font size="4"> Looking at the first big ticket subject of health care: </font>

```{r echo=FALSE}
par(mar=c(11, 4,2,4))
barplot(table(anes_dat$inspre_self),names.arg=c("Refused to Answer","Do nNot Know","No Thought","(1) Gov't Insurance",2,3,"(4) Indifferent",5,6," (7) Private Insurance"),las = 2, main="7pt Scale Of Voters Views Towards Medical Insurance", ylab ="Count")
 mtext("Views", side=1, line =8)
par(mar=c(5.1, 4.1, 4.1, 2.1))

par(mar=c(11, 4,2,4))
barplot(table(anes_dat$presapp_health_x),names.arg=c("Refused to Answer","Do Not Know","Approve Strongly","Approve Not Strongly","Disapprove Not Strongly","Strongly Disapprove"),las = 2, main="Views Towards Obama Care")
 mtext("Views", side=1, line =8)
par(mar=c(5.1, 4.1, 4.1, 2.1))



```



Looking at the '7pt Scale Of Voters Views Towards Medical Insurance' graph, overall voters are spread out on the spectrum of for government insurance plan to private insurance plan. Inface a large number of voters are indifferent towards any specific plan. Furthermore, looking at "Views Towards Obama Care' graph a significant amount of people are in the catorgires of Approve then Disapprove. Majority of the voters are in favor of Obema Care. Obema care is a "law enacted to ensure that all Americans have access to affordable health insurance (abcnews.com)." It is neither forces individuals to get a private insurance plan no government insurance plan. It is just there to help with the primeum coverage and make health insurance more affordable. Obama's views of health care align with the voters because his Care Act works with both types of insurance plans making his views favoriable to all. 

```{r echo=FALSE}
par(mar=c(11, 4,4,4))
barplot(table(anes_dat$inspre_rpc),names.arg=c("Refused to Answer","Do Not Know","(1) Govt Insurance",2,3,"(4) Indifferent",5,6," (7) Private insurance"),las=2, main="7pt Scale Of What The Respondents Think The Republican Party Stands For", ylab ="count")
 mtext("Views", side=1, line =8)
par(mar=c(5.1, 4.1, 4.1, 2.1))
```

Whereas if you look at '7pt Scale Of What The Respondents Think The Republican Party Stands For' graph, most voters think the republican candidate stanads for private insurance plan. This causes him to loose the votes of those that are in favor of government insurance plan (as you can see in the first graph that is close to a thirs of the respondents).

<font size="4"> Looking at the second big ticket subject of handle economic problems: </font>


```{r echo=FALSE, warning=FALSE}
library("knitr")
Economy <- (table(anes_dat$ptyperf_ptyecon))
names(Economy) = c("Refused to answer","Don't Know","Obama","Romney","Both","Neither Candidate")
Economy <- data.frame(Economy)
colnames(Economy) = c("Candidate", "Count")
kable(Economy, caption = "Opinion on Handling of Economy")
```

Based on the table majority of the respondents are in agreement with Obama's handling of the economy. There are 2030 respondents that believe in both candidates stances on how economic problems should be solved but, since they are unclear on one specific candidate. this could should be taken into consideration but should not be counted towards a specific candidate. Hence, unlike the health care subject people seem to be in alignment with Obama on the subject of the Economy. This could be what swung the votes towards Obama despite his approval rating.

<font size="4"> Looking at the third big ticket subject of handling foreign relations with other countries: </font>

```{r echo=FALSE}
par(mar=c(9, 4,2,4))
barplot(table(anes_dat$factor_fpol ),names.arg=c("Refused to Answer","Do Not Know","Incomplete Answer","Missing Answer","Obama", "Romney", "Both"),las=2, main="Foreign Relations With Other Countries", ylab ="count")
 mtext("Views", side=1, line =8)
par(mar=c(5.1, 4.1, 4.1, 2.1))
```

Looking at the 'Foreign Relations with Other Countries' graph, more respondents are in alignment with Obama’s stance on Foreign Relations then they are with Romney's stance. The counts could have been higher but, about 2000 of the respondents have no specific preference towards one candidates’ views. Thus, not taking them in to consideration, 2000 of the Respondents (a significant amount) support Obama’s decisions on foreign policies.


# 5. Conclusion:
As you can see from the analysis above, respondents were in favor of Obama for the next president because majority were in agreement with all of his stances (3 big ticket topics). Majority of the respondents agree with his stance on health care, on foreign affairs and agree that he has handled economic problems well. Hence, even though Obamas approval ratings were low during his reelection campaign, majority of the voters seemed to agree with how he was running the country.


# 6. Biases:
Below mentioned biases should be taken into consideration when looking at the outcome of this study and the outcomes of the survey conducted by ANES.
People were paid to take the surveys. So, they may not cared about answering truthfully. This was the first time ANES used the Internet to conduct its survey ie. the survey was administered via the internet for the first time. Hence, there could have been glitches that resulted in unintended answers. Also, the participants that filled out the online survey were selected from the 'GfK Knowledge Networks'. I am are not sure how DfK Knowledge Networks selects individuals. Furthermore, the survey was also administered by a person. So, a change in tone or body language of the administrator could have influenced the respondent to answer a certain way. Also there is a bias of being observed which can lead to the pressure of the respondent changing their answers. Thus, through out this analysis I have considered this data sample as depicting the population of United States. However, this data from `r ncol(anes_dat)` respondents may not depict the population because of all the biases I have listed above.
